### 2016-10-01 manuscript/manuscript.tex
Built corrected proof PDF

\begin{flushleft}
{\Large
\textbf{Ten Simple Rules for Digital Data Storage}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Edmund Hart $^{1 ,\ast}$,
Pauline Barmby $^{2 }$,
David LeBauer $^{3 }$,
François Michonneau $^{4,5 }$,
Sarah Mount $^{6 }$,
Patrick Mulrooney $^{11 }$,
Timothée Poisot $^{7 }$,
Kara H. Woo $^{8 }$,
Naupaka B. Zimmerman $^{9 }$,
Jeffrey Hollister $^{10 }$\\
\bigskip
\bf{1} University of Vermont, Department of Biology, Burlington
\\
\bf{2} University of Western Ontario, Department of Physics and Astronomy
\\
\bf{3} University of Illinois at Urbana-Champaign, National Center for Supercomputing Applications and Institute for
Genomic Biology
\\
\bf{4} University of Florida, iDigBio, Florida Museum of Natural History
\\
\bf{5} University of Florida, Whitney Laboratory for Marine Bioscience
\\
\bf{6} King's College London, Department of Informatics
\\
\bf{7} Université de Montréal, Département de Sciences Biologiques
\\
\bf{8} Washington State University, Center for Environmental Research, Education, and Outreach
\\
\bf{9} University of Arizona, School of Plant Sciences
\\
\bf{10} US Environmental Protection Agency, Atlantic Ecology Division
\\
\bf{11} University of California at San Diego, San Diego Supercomputer Center
\\
\bigskip
 $\ast$ E-mail: edmund.m.hart@gmail.com

@@

Data is the central currency of science, but the nature of scientific
data has changed dramatically with the rapid pace of technology. This
change has led to the development of a wide variety of data formats,
dataset sizes, data complexity, data use cases, and data sharing
practices. Improvements in high throughput DNA sequencing, sustained
institutional support for large sensor networks {[}1,2{]}, and sky
surveys with large-format digital cameras {[}3{]} have created massive
quantities of data. At the same time, the combination of increasingly
diverse research teams {[}4{]} and data aggregation in portals (e.g.~for
biodiversity data, \href{http://gbif.org/}{GBIF.org} or
\href{https://www.idigbio.org/portal/}{iDigBio}) necessitates increased
coordination among data collectors and institutions {[}5,6{]}. As a
consequence, ``data'' can now mean anything from petabytes of
information stored in professionally-maintained databases, through
spreadsheets on a single computer, to hand-written tables in lab
notebooks on shelves. All remain important, but data curation practices
must continue to keep pace with the changes brought about by new forms
of data and new data collection and storage practices.

While much has been written about both the virtues of data sharing
{[}7,8{]} and the best practices to do so {[}9,10{]}, data storage has
received comparatively less attention. Proper storage is a prerequisite
to sharing, and indeed inadequate storage contributes to the phenomenon
of data decay or to ``data entropy'', where data, whether publicly
shared or not, becomes less accessible through time {[}11--14{]}. Best
practices for data storage often begin and end with, ``deposit your data
in a community standard repository.'' This is good advice, especially
considering your data is most likely to be reused if it is available on
a community site. Community repositories can also provide guidance for
best practices. As an example, if you are archiving sequencing data, a
repository such as those run by the National Center for Biotechnology
Information (NCBI) (e.g.~GenBank) provides not only a location for data
archival, but also encourages a set of practices related to consistent
data formatting and the inclusion of appropriate metadata. However, data
storage policies are highly variable between repositories {[}15{]}. A
data management plan utilizing best practices across all stages of the
data life cycle will facilitate transition from local storage to
repository {[}16{]}. Similarly, having such a plan can facilitate
transition from repository to repository if funding runs out or
requirements change. Good storage practices are important even (or
especially) in cases where data may not fit with an existing repository,
where only derived data products (versus raw data) are suitable for
archiving, or in the case where an existing repository may have lax
standards.

This manuscript describes 10 simple rules for digital data storage that
grew out of a long discussion among instructors for the Software and
Data Carpentry initiatives {[}17,18{]}. Software and Data Carpentry
instructors are scientists from diverse backgrounds who have encountered
a variety of data storage challenges and are active in teaching other
scientists best practices for scientific computing and data management.
Thus, this paper represents a distillation of collective experience, and
hopefully will be useful to scientists facing a variety of data storage
challenges. We additionally provide a glossary of common vocabulary for
readers who may not be familiar with particular terms.

\section*{Rule 1: Anticipate how your data will be
used}\label{rule-1-anticipate-how-your-data-will-be-used}
\addcontentsline{toc}{section}{Rule 1: Anticipate how your data will be
used}

One can avoid most of the troubles encountered during the analysis,
management, and release of data by having a clear roadmap of what to
expect \emph{before} data acquisition starts. For instance:

\begin{itemize}
\tightlist
\item
  How will the raw data be received? Are they delivered by a machine or
  software, or typed-in?
\item
  What is the format expected by the software used for analysis?
\item
  Is there a community standard format for this type of data?
\item
  How much data will be collected, and over what period of time?
\end{itemize}

@@

Also key is the ability to estimate the storage volume needed to store
the data, both during and after the analysis. The required strategy will
differ for datasets of varying size. Smaller datasets (e.g.~a few
megabytes in size) can be managed locally with a simple data management
plan, whereas larger datasets (e.g.~gigabytes to petabytes) will in
almost all cases require careful planning and preparation (Rule 10).

@@

\section*{Rule 2: Know your use case}\label{rule-2-know-your-use-case}
\addcontentsline{toc}{section}{Rule 2: Know your use case}

@@

\begin{itemize}
\tightlist
\item
  Should the raw data be archived (Rule 3)?
\item
  Should the data used for analysis be prepared once, or re-generated
  from the raw data each time (and what difference would this choice
  make for storage, computing requirements, and reproducibility)?
\item
  Can manual corrections be avoided in favor of programmatic or
  self-documenting approaches (e.g., Jupyter notebook or R markdown)?
\item
  How will changes to the data be tracked, and where will these tracked
  changes be logged?
\item
  Will the final data be released, and if so, in what format?
\item
  Are there restrictions or privacy concerns associated with the data
  (e.g.~for survey results with personally identifiable information
  (PII), threatened species, or confidential business information)?
\item
  Will institutional validation be required prior to releasing the data?
\item
  Does the funding agency mandate data deposition in a publicly
  available archive, and if so, when, where, and under what license?
\item
  Does the target journal mandate data deposition?
\end{itemize}

None of these questions have universal answers, nor are they the only
questions to ask before starting data acquisition. But knowing the what,
when, and how of \emph{your} use of the data will bring you close to a
reliable roadmap on how to handle data from acquisition through
publication and archival.

\section*{Rule 3: Keep raw data raw}\label{rule-3-keep-raw-data-raw}
\addcontentsline{toc}{section}{Rule 3: Keep raw data raw}

Since analytical and data processing procedures improve or otherwise
change over time, having access to the `raw' (unprocessed) data can
facilitate future re-analysis and analytical reproducibility. As
processing algorithms improve and computational power increases, new
analyses will be enabled that were not possible at the time of the
original work. If only derived data are stored, it can be difficult for
other researchers to confirm analytical results, to assess the validity
of statistical models, or to directly compare findings across studies.

@@

\section*{Rule 4: Store data in open
formats}\label{rule-4-store-data-in-open-formats}
\addcontentsline{toc}{section}{Rule 4: Store data in open formats}

To maximize accessibility and long-term value, it is preferable to store
data in formats whose specifications are freely available. The
appropriate file type will depend on the data being stored (e.g.~numeric
measurements, text, images, video), but the key idea is that accessing
data should not require proprietary software, hardware, or purchase of a
commercial license. Proprietary formats change, maintaining
organizations go out of business, and changes in license fees make
access to data in proprietary formats unaffordable and risky for
end-users. Examples of open data formats include comma-separated values
(CSV) for tabular data, hierarchical data format (HDF) {[}22{]} and
NetCDF {[}23{]} for hierarchically structured scientific data, portable
network graphics (PNG) for images, KML (or other Open Geospatial
Consortium (OGC) format) for spatial data, and extensible markup
language (XML) for documents. Examples of closed formats include DWG for
AutoCAD drawings, Photoshop document (PSD) for bitmap images, Windows
Media Audio (WMA) for audio recording files, and Microsoft Excel (XLS)
for tabular data. Even if day-to-day processing uses closed formats
(e.g., due to software requirements), data being stored for archival
purposes should be stored in open formats. This is generally not
prohibitive; most closed-source software products enable users to export
data to an open format.

Not only should data be stored in an open format, but it should also be
stored in a format that computers can easily use for processing. This is
especially crucial as datasets become larger. Making data easily-usable
is best achieved by using standard data formats that have open
specifications (e.g., CSV, XML, JSON, HDF5), or by using databases. Such
data formats can be handled by a variety of programming languages, as
efficient and well-tested libraries for parsing them are typically
available. These standard data formats also ensure interoperability,
facilitate re-use, and reduce the chances of data loss or mistakes being
introduced during conversion between formats. Examples of
machine-readable open formats that would \emph{not} be easy to process
include data included in the text of a PDF file, or scanned images of
tabular data from a paper source.

\section*{Rule 5: Data should be structured for
analysis}\label{rule-5-data-should-be-structured-for-analysis}
\addcontentsline{toc}{section}{Rule 5: Data should be structured for
analysis}

To take full advantage of data, it can be useful for it to be structured
in a way that makes use, interpretation, and analysis easy. One such
structure for data stores each variable as a column, each observation as
a row, and each type of observational unit as a table (Fig. 1). The
technical term for this structure is `Codd's 3rd normal form', but it
has been made more accessible as theconcept of \emph{tidy} data
{[}24{]}. When data is organized in this way, the duplication of
information is reduced and it is easier to subset or summarize the
dataset to include the variables or observations of interest.

@@

One axiom about the structure of data and code holds that one should
``write code for humans, write data for computers'' {[}25{]}. When data
can be easily imported and manipulated using familiar software, whether
via a scripting language, a spreadsheet, or any other computer program
that can import these common files, data becomes easier to re-use.
Furthermore, having the source code for the software doing the analysis
available provides provenance for how the data is processed and
analyzed. This makes analysis more transparent, since all assumptions
about the structure of the data are implicitly stated in the source
code. This also enables extraction of the analyses performed, their
reproduction, and their modification.

@@

\section*{Rule 6: Data should be uniquely
identifiable}\label{rule-6-data-should-be-uniquely-identifiable}
\addcontentsline{toc}{section}{Rule 6: Data should be uniquely
identifiable}

To aid reproducibility, the data used in a scientific publication should
be uniquely identifiable. Ideally, datasets should have a unique
identifier such as a Digital Object Identifier (DOI), Archival Resource
Key (ARK), or a persistent URL (PURL). An increasing number of online
services, such as \href{http://figshare.com/}{Figshare},
\href{http://zenodo.org}{Zenodo}, or
\href{http://www.dataone.org}{DataOne}, are able to provide these.
Institutional initiatives also exist, and are known to your local
librarians. Some repositories may require specific identifiers and these
could change with time. For instance, NCBI sequence data will in the
future only be identified by ``accession.version'' IDs. The ``GI''
identifiers (in use since 1994) will be retired in late 2016 {[}26{]}.

Even as identifier standards may change over time, datasets can evolve
over time as well. In order to distinguish between different versions of
the same data, each dataset should have a distinct name, which includes
a version identifier. A simple way to do this is to use date stamps as
part of the dataset name. Using the ISO 8601 standard avoids regional
ambiguities: it mandates the date format \texttt{YYYY-MM-DD} (i.e.~from
largest time unit to smallest). For example, the date ``February 1st,
2015'', while written as \texttt{01-02-2015} in the UK and
\texttt{02-01-2015} in the US, is unambiguous (\texttt{2015-02-01})
under this standard.

Semantic versioning is a richer approach to solving the same problem
{[}27{]}. The CellPack datasets are an example of this {[}28{]}. A
semantic version number takes the form: \texttt{Major.Minor.Patch}, e.g.
\texttt{0.2.7}. The \emph{major version} numbers should be incremented
(or bumped) when a dataset scheme has been updated, or some other change
is made that is not compatible with previous versions of the data with
the same major version number. This means that an experiment using
version \texttt{1.0.0} of the dataset may not run on version
\texttt{2.0.0} without changes to the data analysis. The \emph{minor
version} should be bumped when a change has been made which is
compatible with older versions of the data with the same major version.
This means that any analysis that can be performed on version
\texttt{1.0.0} of the data is repeatable with version \texttt{1.1.0} of
the data. For example, adding a new year in a temporal survey will
result in a bump in the minor version. The \emph{patch version} number
is bumped when typos or bugs have been fixed. For example version
\texttt{1.0.1} of a dataset may fix a typo in version \texttt{1.0.0}.

\section*{Rule 7: Link relevant
metadata}\label{rule-7-link-relevant-metadata}
\addcontentsline{toc}{section}{Rule 7: Link relevant metadata}

@@

Metadata should be as comprehensive as possible, using standards and
conventions of a discipline, and should be machine-readable. Metadata
should always accompany a dataset, wherever it is stored, but the best
way to do this depends on the format of the data. Text files can contain
meta-data in well-defined text files such as XML or JSON. Some file
formats are self-documenting; for example NetCDF, HDF5, and many image
file formats allow for embedded metadata {[}22,23{]}. In a relational
database, metadata tables can be clearly labeled and linked to the data.
Ideally, a schema will be provided that also shows the linkages between
data tables and metadata tables. Another--simpler--scenario is a set of
flat (non-hierarchical) text files--in this case a semantically
versioned, compressed archive should be created that includes metadata.

@@

\section*{Rule 8: Adopt the proper privacy
protocols}\label{rule-8-adopt-the-proper-privacy-protocols}
\addcontentsline{toc}{section}{Rule 8: Adopt the proper privacy
protocols}

In datasets where privacy is important, be sure to have a plan in place
to protect data confidentiality. You should consider the different data
stakeholders when developing privacy protocols for your data storage.
These stakeholders include funding agencies, human subjects or entities,
collaborators, and yourself. Both the NSF and NIH have data sharing
policies in their grant guidelines to prevent sharing personally
identifiable information, and to anonymize data on human subjects.

In small datasets, a lookup table (protecting PII by removing it and
replacing it with a unique id that maps to the sensitive data in an
external dataset) is enough to anonymize a minimal amount of personal
information. Hashing techniques are susceptible to a number of attacks,
and all hashed data can eventually be determined. Famously, New York
City officials shared what they thought was anonymized data on cab
drivers and over 173 million cab rides. However, it was quickly
recognized that the city anonymized the data with a simple MD5 hashing
scheme and all 20 GB of data were de-anonymized in a matter of hours
{[}30{]}. This type of error can be prevented by asking a trusted
colleague or security personnel to try to ``crack'' anonymised data
before releasing it publicly. Often the person who has produced the data
is not well placed to check the fine details of their own security
procedures. If possible, the best solution is to remove any sensitive
data that is not required from the dataset prior to distribution.

@@

\section*{Rule 9: Have a systematic backup
scheme}\label{rule-9-have-a-systematic-backup-scheme}
\addcontentsline{toc}{section}{Rule 9: Have a systematic backup scheme}

Every storage medium can fail, and every failure can result in loss of
data. Researchers should therefore back data up at all stages of the
research process. Data stored on local computers or institutional
servers during the collection and analysis phases of a project should be
backed up to other locations to protect against data loss. No backup
system is failsafe (see the stories of the
\href{https://www.insidehighered.com/news/2014/05/16/dedoose-crash-shows-dangers-handing-data-cloud-services}{Dedoose
crash} and the
\href{http://thenextweb.com/media/2012/05/21/how-pixars-toy-story-2-was-deleted-twice-once-by-technology-and-again-for-its-own-good/}{near
deletion of Toy Story 2}), so more than one backup system should be
used. Ideally you should have two on-site copies (such as on a computer,
an external hard drive, or a tape) and one off-site copy (e.g.~cloud
storage){[}35{]}, with care taken to ensure that the off-site copy is as
secure as the on-site copies. Keeping backups in multiple locations
additionally protects against data loss due to theft or natural
disasters.

@@

\section*{Rule 10: The location and method of data storage depend on how
much data you
have}\label{rule-10-the-location-and-method-of-data-storage-depend-on-how-much-data-you-have}
\addcontentsline{toc}{section}{Rule 10: The location and method of data
storage depend on how much data you have}

The storage method you should choose depends on the size and nature of
your data, the cost of storage and access over time, the time it takes
to transfer the data, how the data will be used, and any privacy
concerns. Data is increasingly generated in the range of many terabytes
by environmental sensors, satellites, automated analytical tools,
simulation models, and nucleic acid sequencers. Even larger data
generating machines, like the Large Hadron Collider (LHC) and the Large
Scale Synoptic Survey Telescope (LSST), generate many terabytes (TB) per
day, rapidly accumulating to petabyte (PB) scale over the course of any
particular study. While the cost of storage continues to decrease, the
volume of data to be stored impacts the choice of storage methods and
locations: for large datasets it is necessary to balance the cost of
storage with the time of access and costs of re-generating the data.
With new commercial cloud offerings (e.g., Amazon S3) the cost of
retrieving the data might exceed the cost of analysis or of
re-generating the data from scratch.

When data takes too long to transfer or is costly to store, it can
become more efficient to use a computer system for analysis that can
directly access and use the data in place instead of first transferring
it to a local machine. Inactive data can be put in longer-term storage;
this is less expensive, but can take longer to retrieve. Some storage
systems automatically migrate `stale' files to longer term storage.
Alternatively, some computing can be done `in the database' or `on disk'
via database query languages (e.g.~SQL, MapReduce) that perform basic
arithmetic, or via the use of procedural languages (e.g.~R, Python, C)
embedded in the database server. Modern database technologies such as
HDFS and Spark allow these computations to be done on data of almost any
size. When data is larger than locally available RAM, it can be handled
by conducting analyses on a `big memory' node, which most
high-performance computing centers have deployed. Relying on tight
software/hardware integration, these can allow for the analysis of
datasets around 1-4 TB in size. This allows the user to read in and use
a large dataset without special tools.

>>>>

\begin{flushleft}
{\Large
\textbf{Ten Simple Rules for Digital Data Storage}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Edmund M. Hart $^{1 ,\ast}$,
Pauline Barmby $^{2 }$,
David LeBauer $^{3 }$,
François Michonneau $^{4,5 }$,
Sarah Mount $^{6 }$,
Patrick Mulrooney $^{11 }$,
Timothée Poisot $^{7 }$,
Kara H. Woo $^{8 }$,
Naupaka B. Zimmerman $^{9 }$,
Jeffrey W. Hollister $^{10 }$\\
\bigskip
\bf{1} University of Vermont, Department of Biology, Burlington
\\
\bf{2} University of Western Ontario, Department of Physics and Astronomy
\\
\bf{3} University of Illinois at Urbana-Champaign, National Center for Supercomputing Applications and Institute for
Genomic Biology
\\
\bf{4} University of Florida, iDigBio, Florida Museum of Natural History
\\
\bf{5} University of Florida, Whitney Laboratory for Marine Bioscience
\\
\bf{6} King's College London, Department of Informatics
\\
\bf{7} Université de Montréal, Département de Sciences Biologiques
\\
\bf{8} Washington State University, Center for Environmental Research, Education, and Outreach
\\
\bf{9} University of Arizona, School of Plant Sciences
\\
\bf{10} US Environmental Protection Agency, Atlantic Ecology Division
\\
\bf{11} University of California at San Diego, San Diego Supercomputer Center
\\
\bigskip
 $\ast$ E-mail: edmund.m.hart@gmail.com

@@

Data is the central currency of science, but the nature of scientific
data has changed dramatically with the rapid pace of technology. This
change has led to the development of a wide variety of data formats,
dataset sizes, data complexity, data use cases, and data sharing
practices. Improvements in high-throughput DNA sequencing, sustained
institutional support for large sensor networks {[}1,2{]}, and sky
surveys with large-format digital cameras {[}3{]} have created massive
quantities of data. At the same time, the combination of increasingly
diverse research teams {[}4{]} and data aggregation in portals (e.g.,
for biodiversity data, \href{http://gbif.org/}{GBIF.org} or
\href{https://www.idigbio.org/portal/}{iDigBio}) necessitates increased
coordination among data collectors and institutions {[}5,6{]}. As a
consequence, ``data'' can now mean anything from petabytes of
information stored in professionally maintained databases, to
spreadsheets on a single computer, to handwritten tables in lab
notebooks on shelves. All remain important, but data curation practices
must continue to keep pace with the changes brought about by new forms
of data and new data collection and storage practices.

While much has been written about both the virtues of data sharing
{[}7,8{]} and the best practices to do so {[}9,10{]}, data storage has
received comparatively less attention. Proper storage is a prerequisite
to sharing, and indeed inadequate storage contributes to the phenomenon
of data decay or to ``data entropy,'' in which data, whether publicly
shared or not, becomes less accessible through time {[}11--14{]}. Best
practices for data storage often begin and end with this statement:
``Deposit your data in a community standard repository.'' This is good
advice, especially considering your data is most likely to be reused if
it is available on a community site. Community repositories can also
provide guidance for best practices. As an example, if you are archiving
sequencing data, a repository such as those run by the National Center
for Biotechnology Information (NCBI) (e.g., GenBank) not only provides a
location for data archival, but also encourages a set of practices
related to consistent data formatting and the inclusion of appropriate
metadata. However, data storage policies are highly variable between
repositories {[}15{]}. A data management plan utilizing best practices
across all stages of the data life cycle will facilitate transition from
local storage to repository {[}16{]}. Similarly, having such a plan can
facilitate transition from repository to repository if funding runs out
or requirements change. Good storage practices are important even (or
especially) in cases when data may not fit with an existing repository,
when only derived data products (versus raw data) are suitable for
archiving, or when an existing repository may have lax standards.

This article describes ten simple rules for digital data storage that
grew out of a long discussion among instructors for the Software and
Data Carpentry initiatives {[}17,18{]}. Software and Data Carpentry
instructors are scientists from diverse backgrounds who have encountered
a variety of data storage challenges and are active in teaching other
scientists best practices for scientific computing and data management.
Thus, this paper represents a distillation of collective experience, and
hopefully will be useful to scientists facing a variety of data storage
challenges. We additionally provide a glossary of common vocabulary for
readers who may not be familiar with particular terms.

\section*{Rule 1: Anticipate How Your Data Will Be
Used}\label{rule-1-anticipate-how-your-data-will-be-used}
\addcontentsline{toc}{section}{Rule 1: Anticipate How Your Data Will Be
Used}

One can avoid most of the troubles encountered during the analysis,
management, and release of data by having a clear roadmap of what to
expect before data acquisition starts. For instance:

\begin{itemize}
\tightlist
\item
  How will the raw data be received? Are they delivered by a machine or
  software, or typed in?
\item
  What is the format expected by the software used for analysis?
\item
  Is there a community standard format for this type of data?
\item
  How much data will be collected, and over what period of time?
\end{itemize}

@@

Also key is the ability to estimate the storage volume needed to store
the data, both during and after the analysis. The required strategy will
differ for datasets of varying size. Smaller datasets (e.g., a few
megabytes in size) can be managed locally with a simple data management
plan, whereas larger datasets (e.g., gigabytes to petabytes) will in
almost all cases require careful planning and preparation (Rule 10).

@@

\section*{Rule 2: Know Your Use Case}\label{rule-2-know-your-use-case}
\addcontentsline{toc}{section}{Rule 2: Know Your Use Case}

@@

\begin{itemize}
\tightlist
\item
  Should the raw data be archived (Rule 3)?
\item
  Should the data used for analysis be prepared once, or re-generated
  from the raw data each time (and what difference would this choice
  make for storage, computing requirements, and reproducibility)?
\item
  Can manual corrections be avoided in favor of programmatic or
  self-documenting approaches (e.g., Jupyter notebook or R markdown)?
\item
  How will changes to the data be tracked, and where will these tracked
  changes be logged?
\item
  Will the final data be released, and if so, in what format?
\item
  Are there restrictions or privacy concerns associated with the data
  (e.g., survey results with personally identifiable information
  {[}PII{]}, threatened species, or confidential business information)?
\item
  Will institutional validation be required prior to releasing the data?
\item
  Does the funding agency mandate data deposition in a publicly
  available archive, and if so, when, where, and under what license?
\item
  Does the target journal mandate data deposition?
\end{itemize}

None of these questions have universal answers, nor are they the only
questions to ask before starting data acquisition. But knowing the what,
when, and how of your use of the data will bring you close to a reliable
roadmap on how to handle data from acquisition through publication and
archival.

\section*{Rule 3: Keep Raw Data Raw}\label{rule-3-keep-raw-data-raw}
\addcontentsline{toc}{section}{Rule 3: Keep Raw Data Raw}

Since analytical and data processing procedures improve or otherwise
change over time, having access to the ``raw'' (unprocessed) data can
facilitate future re-analysis and analytical reproducibility. As
processing algorithms improve and computational power increases, new
analyses will be enabled that were not possible at the time of the
original work. If only derived data are stored, it can be difficult for
other researchers to confirm analytical results, to assess the validity
of statistical models, or to directly compare findings across studies.

@@

\section*{Rule 4: Store Data in Open
Formats}\label{rule-4-store-data-in-open-formats}
\addcontentsline{toc}{section}{Rule 4: Store Data in Open Formats}

To maximize accessibility and long-term value, it is preferable to store
data in formats that have freely available specifications. The
appropriate file type will depend on the data being stored (e.g.,
numeric measurements, text, images, video), but the key idea is that
accessing data should not require proprietary software, hardware, or
purchase of a commercial license. Proprietary formats change,
maintaining organizations go out of business, and changes in license
fees make access to data in proprietary formats unaffordable and risky
for end-users. Examples of open data formats include comma-separated
values (CSV) for tabular data, hierarchical data format (HDF) {[}22{]}
and NetCDF {[}23{]} for hierarchically structured scientific data,
portable network graphics (PNG) for images, KML (or other Open
Geospatial Consortium {[}OGC{]} format) for spatial data, and extensible
markup language (XML) for documents. Examples of closed formats include
DWG for AutoCAD drawings, Photoshop document (PSD) for bitmap images,
Windows Media Audio (WMA) for audio recording files, and Microsoft Excel
(XLS) for tabular data. Even if day-to-day processing uses closed
formats (e.g., due to software requirements), data being stored for
archival purposes should be stored in open formats. This is generally
not prohibitive; most closed-source software products enable users to
export data to an open format.

Not only should data be stored in an open format, but it should also be
stored in a format that computers can easily use for processing. This is
especially crucial as datasets become larger. Making data easily usable
is best achieved by using standard data formats that have open
specifications (e.g., CSV, XML, JSON, HDF5), or by using databases. Such
data formats can be handled by a variety of programming languages, as
efficient and well-tested libraries for parsing them are typically
available. These standard data formats also ensure interoperability,
facilitate re-use, and reduce the chances of data loss or mistakes being
introduced during conversion between formats. Examples of
machine-readable open formats that would not be easy to process include
data included in the text of a PDF file, or scanned images of tabular
data from a paper source.

\section*{Rule 5: Data Should Be Structured for
Analysis}\label{rule-5-data-should-be-structured-for-analysis}
\addcontentsline{toc}{section}{Rule 5: Data Should Be Structured for
Analysis}

To take full advantage of data, it can be useful for it to be structured
in a way that makes use, interpretation, and analysis easy. One such
structure for data stores each variable as a column, each observation as
a row, and each type of observational unit as a table (Fig. 1). The
technical term for this structure is ``Codd's 3rd normal form'', but it
has been made more accessible as the concept of tidy data {[}24{]}. When
data is organized in this way, the duplication of information is reduced
and it is easier to subset or summarize the dataset to include the
variables or observations of interest.

@@

One axiom about the structure of data and code holds that one should
``write code for humans, write data for computers'' {[}25{]}. When data
can be easily imported and manipulated using familiar software (whether
via a scripting language, a spreadsheet, or any other computer program
that can import these common files), data becomes easier to re-use.
Furthermore, having the source code for the software doing the analysis
available provides provenance for how the data is processed and
analyzed. This makes analysis more transparent, since all assumptions
about the structure of the data are implicitly stated in the source
code. This also enables extraction of the analyses performed, their
reproduction, and their modification.

@@

\section*{Rule 6: Data Should Be Uniquely
Identifiable}\label{rule-6-data-should-be-uniquely-identifiable}
\addcontentsline{toc}{section}{Rule 6: Data Should Be Uniquely
Identifiable}

To aid reproducibility, the data used in a scientific publication should
be uniquely identifiable. Ideally, datasets should have a unique
identifier such as a Digital Object Identifier (DOI), Archival Resource
Key (ARK), or a persistent URL (PURL). An increasing number of online
services, such as \href{http://figshare.com/}{Figshare},
\href{http://zenodo.org}{Zenodo}, or
\href{http://www.dataone.org}{DataOne}, are able to provide these.
Institutional initiatives also exist and are known to your local
librarians. Some repositories may require specific identifiers, and
these could change with time. For instance, NCBI sequence data will in
the future only be identified by ``accession.version'' IDs. The ``GI''
identifiers (in use since 1994) will be retired in late 2016 {[}26{]}.

Even as identifier standards may change over time, datasets can evolve
over time as well. In order to distinguish between different versions of
the same data, each dataset should have a distinct name, which includes
a version identifier. A simple way to do this is to use date stamps as
part of the dataset name. Using the ISO 8601 standard avoids regional
ambiguities: it mandates the date format \texttt{YYYY-MM-DD} (i.e.~from
largest time unit to smallest). For example, the date ``February 1,
2015,'' while written as \texttt{01-02-2015} in the UK and
\texttt{02-01-2015} in the US, is unambiguous (\texttt{2015-02-01})
under this standard.

Semantic versioning is a richer approach to solving the same problem
{[}27{]}. The CellPack datasets are an example of this {[}28{]}. A
semantic version number takes the form: \texttt{Major.Minor.Patch} e.g.,
\texttt{0.2.7}. The \emph{major version} numbers should be incremented
(or bumped) when a dataset scheme has been updated, or some other change
is made that is not compatible with previous versions of the data with
the same major version number. This means that an experiment using
version \texttt{1.0.0} of the dataset may not run on version
\texttt{2.0.0} without changes to the data analysis. The \emph{minor
version} should be bumped when a change has been made that is compatible
with older versions of the data with the same major version. This means
that any analysis that can be performed on version \texttt{1.0.0} of the
data is repeatable with version \texttt{1.1.0} of the data. For example,
adding a new year in a temporal survey will result in a bump in the
minor version. The \emph{patch version} number is bumped when typos or
bugs have been fixed. For example version \texttt{1.0.1} of a dataset
may fix a typo in version \texttt{1.0.0}.

\section*{Rule 7: Link Relevant
Metadata}\label{rule-7-link-relevant-metadata}
\addcontentsline{toc}{section}{Rule 7: Link Relevant Metadata}

@@

Metadata should be as comprehensive as possible, using standards and
conventions of a discipline, and should be machine-readable. Metadata
should always accompany a dataset, wherever it is stored, but the best
way to do this depends on the format of the data. Text files can contain
metadata in well-defined text files such as XML or JSON. Some file
formats are self-documenting; for example, NetCDF, HDF5, and many image
file formats allow for embedded metadata {[}22,23{]}. In a relational
database, metadata tables can be clearly labeled and linked to the data.
Ideally, a schema will be provided that also shows the linkages between
data tables and metadata tables. Another---simpler---scenario is a set
of flat (non-hierarchical) text files---in this case a semantically
versioned, compressed archive should be created that includes metadata.

@@

\section*{Rule 8: Adopt the Proper Privacy
Protocols}\label{rule-8-adopt-the-proper-privacy-protocols}
\addcontentsline{toc}{section}{Rule 8: Adopt the Proper Privacy
Protocols}

In datasets for which privacy is important, be sure to have a plan in
place to protect data confidentiality. You should consider the different
data stakeholders when developing privacy protocols for your data
storage. These stakeholders include funding agencies, human subjects or
entities, collaborators, and yourself. Both the National Science
Foundation (NSF) and National Institutes of Health (NIH) have data
sharing policies in their grant guidelines to prevent sharing personally
identifiable information, and to anonymize data on human subjects.

In small datasets, a lookup table (protecting PII by removing it and
replacing it with a unique ID that maps to the sensitive data in an
external dataset) is enough to anonymize a minimal amount of personal
information. Hashing techniques are susceptible to a number of attacks,
and all hashed data can eventually be determined. Famously, New York
City officials shared what they thought was anonymized data on cab
drivers and over 173 million cab rides. However, it was quickly
recognized that the city anonymized the data with a simple MD5 hashing
scheme and all 20 GB of data were de-anonymized in a matter of hours
{[}30{]}. This type of error can be prevented by asking a trusted
colleague or security personnel to try to ``crack'' anonymised data
before releasing it publicly. Often the person who has produced the data
is not well placed to check the fine details of their own security
procedures. If possible, the best solution is to remove any sensitive
data that is not required from the dataset prior to distribution.

@@

\section*{Rule 9: Have a Systematic Backup
Scheme}\label{rule-9-have-a-systematic-backup-scheme}
\addcontentsline{toc}{section}{Rule 9: Have a Systematic Backup Scheme}

Every storage medium can fail, and every failure can result in loss of
data. Researchers should therefore back up data at all stages of the
research process. Data stored on local computers or institutional
servers during the collection and analysis phases of a project should be
backed up to other locations to protect against data loss. No backup
system is failsafe (see the stories of the
\href{https://www.insidehighered.com/news/2014/05/16/dedoose-crash-shows-dangers-handing-data-cloud-services}{Dedoose
crash} and the
\href{http://thenextweb.com/media/2012/05/21/how-pixars-toy-story-2-was-deleted-twice-once-by-technology-and-again-for-its-own-good/}{near
deletion of Toy Story 2}), so more than one backup system should be
used. Ideally you should have two on-site copies (such as on a computer,
an external hard drive, or a tape) and one off-site copy (e.g., cloud
storage){[}35{]}, with care taken to ensure that the off-site copy is as
secure as the on-site copies. Keeping backups in multiple locations
additionally protects against data loss due to theft or natural
disasters.

@@

\section*{Rule 10: The Location and Method of Data Storage Depend on How
Much Data You
Have}\label{rule-10-the-location-and-method-of-data-storage-depend-on-how-much-data-you-have}
\addcontentsline{toc}{section}{Rule 10: The Location and Method of Data
Storage Depend on How Much Data You Have}

The storage method you should choose depends on the size and nature of
your data, the cost of storage and access over time, the time it takes
to transfer the data, how the data will be used, and any privacy
concerns. Data is increasingly generated in the range of many terabytes
by environmental sensors, satellites, automated analytical tools,
simulation models, and nucleic acid sequencers. Even larger
data-generating machines, like the Large Hadron Collider (LHC) and the
Large Scale Synoptic Survey Telescope (LSST), generate many TB per day,
rapidly accumulating to petabyte (PB) scale over the course of any
particular study. While the cost of storage continues to decrease, the
volume of data to be stored impacts the choice of storage methods and
locations: for large datasets it is necessary to balance the cost of
storage with the time of access and costs of re-generating the data.
With new commercial cloud offerings (e.g., Amazon S3) the cost of
retrieving the data might exceed the cost of analysis or of
re-generating the data from scratch.

When data takes too long to transfer or is costly to store, it can
become more efficient to use a computer system for analysis that can
directly access and use the data in place instead of first transferring
it to a local machine. Inactive data can be put in longer-term storage;
this is less expensive, but can take longer to retrieve. Some storage
systems automatically migrate ``stale'' files to longer-term storage.
Alternatively, some computing can be done ``in the database'' or ``on
disk'' via database query languages (e.g., SQL, MapReduce) that perform
basic arithmetic, or via the use of procedural languages (e.g., R,
Python, C) embedded in the database server. Modern database technologies
such as HDFS and Spark allow these computations to be done on data of
almost any size. When data is larger than locally available RAM, it can
be handled by conducting analyses on a ``big memory'' node, which most
high-performance computing centers have deployed. Relying on tight
software/hardware integration, these can allow for the analysis of
datasets around 1--4 TB in size. This allows the user to read in and use
a large dataset without special tools.
